---
title: 'Shoegaze: from the 90s to the present'
author: "Vincent Huf"
date: "9-2-2021"
output: html_document
---


```{r}
library(tidyverse)
library(spotifyr)

```

```{r}
just <- get_playlist_audio_features("", "6NqUXlpEEFZNytZiGLi4kW")
slowdive <- get_playlist_audio_features("", "12KhHp6mG8xZz1urkeBhoS")


combi <-
  bind_rows(
    just %>% mutate(category = "90s"),
    slowdive %>% mutate(category = "2018")
  )
```

```{r}
combi %>%
  ggplot(aes(x = loudness, y = instrumentalness, color = mode, size = tempo)) + geom_point(alpha = 0.5) + 
  
  facet_wrap(~category) + ggtitle("Slowdive in the 90s vs the present")+ theme(plot.title = element_text(size = 15))+theme_light() +      
  labs(               
    x = "Loudness",
    y = "Instrumentalness",
    colour = "Mode",
    size = "Tempo") + scale_color_binned(n.breaks = 3, show.limits = TRUE) + scale_size_continuous() + scale_x_continuous() + scale_y_continuous()
```

\newpage

This graph compares slowdives music from the 90s, and their reunion album from 2018.

The general conclusion from this comparison is that the newer shoegaze songs from slowdive are louder than the songs from the 90s. Also, a lot of songs in the 90s are clustered together, which indicates some kind of sweet spot for loudness and instrumentalism. Last of all, the newer songs tend to be in major key. In the 90s, shoegaze was considered as 'emo' music, which might explain why there were more songs in minor key back then. It would be interesting to see if this change from minor to major also appears in other bands.

The two outliers with very low intrumentalism are two songs which also have a very low speechiness. I'm not sure yet why, because this seems to contradict. It might be because their vocals really blend into the guitar, and this might be harder to differentiate for the algorithm.  